An elaborate data set on human gait and the effect of mechanical perturbations

### 2014-12-12 paper.tex
Update paper.tex
Edits by Ton.

  Here we share a rich gait data set collected from fifteen subjects walking at
  three speeds on an instrumented treadmill. Each trial consists of 120 seconds
  of normal walking and 480 seconds of walking while being longitudinally
  and/or laterally perturbed during each stance phase with pseudo-random
  fluctuations in the speed of the treadmill belt. A total of approximately 1.5
  hours of normal walking ($>5000$ gait cycles) and 6 hours of perturbed walking
  ($>20,000$ gait cycles) is included in the data set. We provide full body
  marker trajectories and ground reaction loads in addition to a presentation of
  processed data that includes gait landmarks, 2D joint angles, angular rates,
  and joint torques and the software used for the computations. The protocol is
  described in detail along with additional elaborate meta data. This data can
  likely be useful for validating or generating mathematical models that are
  capable of simulating normal periodic gait and non-periodic, perturbed gaits.

@@

\section*{Introduction}
%
The collection of dynamical data during human walking has a long history
beginning with the first motion pictures and now with modern marker based
motion capture techniques, image and video analysis, and high fidelity ground
reaction load measurements. Even though years of data on thousands of subjects
now exist, this data is not widely disseminated, well organized, nor available
with few or no restrictions. David Winter's published normative gait
data~\cite{Winter1990} is widely used in biomechanics studies, yet it comes
from relatively few subjects and only a small number of gait cycles per
subject. This small source has successfully inspired powered prosthetic control
design~\cite{Sup2008}, but success in other fields using large sets of
data for discovery lead one to believe that may be possible in the
biomechanical field too. To find out if that hypothesis is true, our data needs
to be shared extensively, organized, and curated to enabled analysts.

@@

Chester et. al~\cite{Chester2007} report on a large gait database comparison
where one database contained kinematic data of 409 gait cycles of children from
1 to 7 years old but the data does not seem to be publicly available. Tirosh
et. al, recognized the need of a comprehensive data base for clinical gait data
and created the Gaitabase~\cite{Tirosh2010}.  This database may contain a
substantial amount of data but it is encumbered by a very complicated and
restrictive license and sharing scheme. The University of Wisconsin at LaCrosse
has a normative gait data set~\cite{Wilson} from 25 subjects with lower
extremity marker data and force plate measurements from a single stance phase.

More recent examples of biomechanist's sharing their data alongside
publications are: \cite{Bogert2013} which includes processed joint kinematics
from eleven subjects walking for few gait cycles and \cite{Wang2014} who
includes data from 10 subjects walking for five minutes each at five different
speeds but only a small set of lower extremity markers are present. The second
is notable because it publishes the data in Dryad, a modern citable data
repository.

@@

Furthermore, we include a small Apache licensed open source software library
for basic gait analysis and demonstrate it's use in the paper. The combination
of the open data and open software allow the results presented within to be
computationally reproducible and instructions are included in the associated
repository for doing so.

\section*{Participants}
%
Fifteen able bodied subjects including four females and eleven males with an
average age of $24\pm4$ years, height of $1.75\pm0.09$ m, mass of $74\pm13$ kg
participated in the study. The data has been anonymized with respect to the
participants' identities and an unique identification number was assigned to
each subject. A selection of the meta data collected for each subject is shown
in Table~\ref{tab:subjects}.
%
\begin{table*}
  \cprotect\caption{Information about the 15 participants. The final three
    columns give the trial numbers associated with each nominal treadmill
    speed. Additional trials with subject 0 are trials with no subject, i.e.
    unloaded trials that can be used for force compensation purposes, and are
    not shown in the table. Generated by \verb|src/subject_table.py|.}
  \centering
  \input{tables/subjects.tex}
  \label{tab:subjects}
\end{table*}

\section*{Equipment}
%
The Human Motion and Control Lab at Cleveland State University has modern human
motion measurement equipment. Our system includes:

@@

Cortex alone is capable of delivering data from the cameras along with force
plates, and analog sensors (EMG/Accelerometer) through a National Instruments
USB-6255 data acquisition unit. D-Flow is required to collect data from any
digital sensors and the treadmill's motion (lateral, pitch, and belts). D-Flow
can output multiple types of files which contain the different data.

@@

After a briefing by the experimentalist on the trial protocol the subject
follows the verbal instructions of the experimentalist and the on-screen
instructions from the video display in front of the subject. The protocol goes
as so:
%
\begin{enumerate}
  \item Subject steps onto the treadmill and markers are identified with
    Cortex.
  \item The safety cable is attached loosely to the rock climbing harness such
    that no undue forces are acting on the subject during walking, but that the
    harness will prevent a full fall.
  \item The subject starts by stepping on sides of treadmill so that feet don't
    touch the force plates and the force plate signals are zeroed.
  \item Once notified by the screen, the subject stands in the initialization
    pose: standing straight up, looking forward, arms out by their sides (~45
    degrees) and the event is manually recorded by the operator.
  \item A countdown to the first normal walking phase is displayed. At the end
    of the countdown the treadmill starts at the specified speed and the
    subject is instructed to walk normally, to focus on the road in the
    screen, and not to look at their feet.
  \item After 1 minute of normal walking, the longitudinal perturbation phase
    begins.
  \item After 8 minutes of walking under the influence of the perturbations,
    the normal walking phase begins.
  \item After 1 minute, a countdown is shown and the treadmill decelerates to a
    stop.
  \item The subject is instructed to step off of the force plates for 10
    seconds.
  \item The subject could take a break before each additional trial.
\end{enumerate}

@@

The \verb|GaitData| class is then used to compute things such as gait landmarks
(toe off and heel strike times), basic 2D inverse dynamics, and to store the
data into a Pandas \verb|Panel| with each gait cycle on the main axis at a
specified sample rate. This data is also saved to disk in HDF5 format.

@@

\section*{Data Limitations}
%
The data is provided in good faith with great attention to detail. The
following list gives various notes and warnings about the data that should be
taken into account when making use of it.
%
\begin{itemize}
  \item Be sure to read the notes in each meta data file for details about
    possible anomalies in that particular trial.
  \item The subject identification number 0 stands for "no subject" and was
    used whenever data was collected from the system with no subject on the
    treadmill, for example during the trials that were intended to be used for
    inertial compensation purposes if the treadmill was moving. These trials
    play through the exact protocol as those with a human subject and the
    matching trials are indicated in the meta data.
  \item Trials 1 and 2 were not recorded as part of this study. Those trial
    identification numbers were reserved for early data exploration from data
    collected in other studies.
  \item Trials 37, 38, and 39 do not exist. The numbers were accidentally
    skipped.
  \item Trials 9, 10, and 11 used a slightly different event definition where the
    calibration poses were not explicitly tagged by an event, yet the protocol
    was the identical to the following trials.
  \item During trials 9-15 we used wooden blocks to fix the treadmill to the
    concrete floor to eliminate the treadmill's lowest factory vibration mode
    (~5 hz). But these blocks seem to have corrupted the force plate
    measurements by imposing undue stresses on the system. The force plate
    measurements should not be used from these trials, but the marker data is
    fine.
  \item We did not record unloaded compensation trials for trials 9-15.
    Regardless, they would be useless due to the previous point.
  \item Trials 6-8 use an early experimental protocol which divided the
    perturbation sections into three sections: longitudinal perturbations,
    lateral perturbations, and a combination of each. We then learned the
    treadmill had a low vibrational mode which significantly affects the force
    plate measurements. The force measurements during these trials are
    corrupted by this vibrational mode and should be used with caution or not
    at all.
  \item Trials 6-8 use a XX lower body marker set only.
  \item The ankle joint torques computed from subject 9's data in trials 25-27
    are abnormal and should be used with caution or not at all. We were not
    about to locate the source of the error, but it is likely related to the
    force calibration.
\end{itemize}

>>>>

  Here we share a rich gait data set collected from fifteen subjects walking at
  three speeds on an instrumented treadmill. Each trial consists of 120 seconds
  of normal walking and 480 seconds of walking while being longitudinally
  perturbed during each stance phase with pseudo-random
  fluctuations in the speed of the treadmill belt. A total of approximately 1.5
  hours of normal walking ($>5000$ gait cycles) and 6 hours of perturbed walking
  ($>20,000$ gait cycles) is included in the data set. We provide full body
  marker trajectories and ground reaction loads in addition to a presentation of
  processed data that includes gait events, 2D joint angles, angular rates,
  and joint torques and the software used for the computations. The protocol is
  described in detail along with additional elaborate meta data. This data can
  likely be useful for validating or generating mathematical models that are
  capable of simulating normal periodic gait and non-periodic, perturbed gaits.

@@

\section*{Introduction}
%
The collection of dynamical data during human walking has a long history
beginning with the first motion pictures and now with modern marker based
motion capture techniques, image and video analysis, and high fidelity ground
reaction load measurements. Even though years of data on thousands of subjects
now exist, this data is not widely disseminated, well organized, nor available
with few or no restrictions. David Winter's published normative gait
data~\cite{Winter1990} is widely used in biomechanics studies, yet it comes
from relatively few subjects and only a small number of gait cycles per
subject. This small source has successfully inspired powered prosthetic control
design~\cite{Sup2008}, but success in other fields using large sets of
data for discovery lead one to believe that more elaborate data sets may 
benefit the field of human motion studies.
To enable such work, data needs
to be shared extensively, organized, and curated to enabled analysts.

@@

Chester et. al~\cite{Chester2007} report on a large gait database comparison
where one database contained kinematic data of 409 gait cycles of children from
1 to 7 years old but the data does not seem to be publicly available. Tirosh
et. al, recognized the need of a comprehensive data base for clinical gait data
and created the Gaitabase~\cite{Tirosh2010}.  This database may contain a
substantial amount of data but it is encumbered by a very complicated and
restrictive license and sharing scheme. The University of Wisconsin at LaCrosse
has a normative gait data set~\cite{Wilson} from 25 subjects with lower
extremity marker data and force plate measurements from a single stance phase (TvdB: gait cycle?).

More recent examples of biomechanists sharing their data alongside
publications are: \cite{Bogert2013} which includes processed joint kinematics
from eleven subjects walking for few gait cycles and \cite{Wang2014} who
includes data from 10 subjects walking for five minutes each at five different
speeds but only a small set of lower extremity markers are present. The second
is notable because it publishes the data in Dryad, a modern citable data
repository.

@@

Furthermore, we include a small Apache licensed open source software library
for basic gait analysis and demonstrate its use in the paper. The combination
of the open data and open software allow the results presented within to be
computationally reproducible and instructions are included in the associated
repository for doing so.

\section*{Participants}
The study was approved by the Institutional Review Board of Cleveland State University
and informed consent was obtained from all participants.
%
Fifteen able bodied subjects including four females and eleven males with an
average age of $24\pm4$ years, height of $1.75\pm0.09$ m, mass of $74\pm13$ kg
participated in the study. The data has been anonymized with respect to the
participants' identities and a unique identification number was assigned to
each subject. A selection of the meta data collected for each subject is shown
in Table~\ref{tab:subjects}.
%
\begin{table*}
  \cprotect\caption{Information about the 15 participants. The final three
    columns give the trial numbers associated with each nominal treadmill
    speed. Additional trials with subject 0 are trials with no subject, i.e.
    unloaded trials that can be used for inertial compensation purposes, and are
    not shown in the table. Generated by \verb|src/subject_table.py|.}
  \centering
  \input{tables/subjects.tex}
  \label{tab:subjects}
\end{table*}

\section*{Equipment}
%
The data were collected in the Laboratory for Human Motion and Control at Cleveland State University,
using the following equipment:

@@

The Cortex software delivers 3D marker trajectories from the cameras along with data
from force
plates and analog sensors (EMG/Accelerometer) through a National Instruments
USB-6255 data acquisition unit. D-Flow is required to collect data from any
digital sensors and to control the treadmill's motion (lateral, pitch, and belts). D-Flow
can process the data in real time and/or export data to file.

@@

After obtaining informed consent and a briefing by the experimentalist on the trial protocol, the subject
follows the verbal instructions of the experimentalist and the on-screen
instructions from the video display in front of the subject. The protocol goes
as follows:
%
\begin{enumerate}
  \item Subject steps onto the treadmill and markers are identified with
    Cortex.
  \item The safety cable is attached loosely to the rock climbing harness such
    that no undue forces are acting on the subject during walking, but that the
    harness will prevent a full fall.
  \item The subject starts by stepping on sides of treadmill so that feet don't
    touch the force plates and the force plate signals are zeroed.
  \item Once notified by the screen, the subject stands in the initialization
    pose: standing straight up, looking forward, arms out by their sides (~45
    degrees) and the event is manually recorded by the operator.
  \item A countdown to the first normal walking phase is displayed. At the end
    of the countdown the treadmill starts at the specified speed and the
    subject is instructed to walk normally, to focus on the road in the
    screen, and not to look at their feet.
  \item After 1 minute of normal walking, the longitudinal perturbation phase
    begins.
  \item After 8 minutes of walking under the influence of the perturbations,
    the normal walking phase begins.
  \item After 1 minute, a countdown is shown and the treadmill decelerates to a
    stop.
  \item The subject is instructed to step off of the force plates for 10
    seconds.
  \item The subject could take a break before each additional trial.
\end{enumerate}

@@

The \verb|GaitData| class is then used to compute things such as gait landmarks
(toe off and heel strike times), basic 2D kinematics and inverse dynamics, and to store the
data into a Pandas \verb|Panel| with each gait cycle on the main axis at a
specified sample rate. This data is also saved to disk in HDF5 format.

@@

\section*{Data Limitations}
%
The data is provided in good faith with great attention to detail. The
following list gives various notes and warnings about the data that should be
taken into account when making use of it.
%
\begin{itemize}
  \item Be sure to read the notes in each meta data file for details about
    possible anomalies in that particular trial.
  \item The subject identification number 0 stands for "no subject" and was
    used whenever data was collected from the system with no subject on the
    treadmill, for example during the trials that were intended to be used for
    inertial compensation purposes if the treadmill was moving. These trials
    play through the exact protocol as those with a human subject and the
    matching trials are indicated in the meta data. (TvdB: can the user tell
    which compensation file to use with which subject files?  By the date on
    the file?)
  \item Trials 1 and 2 were not recorded as part of this study. Those trial
    identification numbers were reserved for early data exploration from data
    collected in other studies.
  \item Trials 37, 38, and 39 do not exist. The numbers were accidentally
    skipped.
  \item Trials 9, 10, and 11 used a slightly different event definition where the
    calibration poses were not explicitly tagged by an event, yet the protocol
    was the identical to the following trials.
  \item During trials 9-15 we used wooden blocks to fix the treadmill to the
    concrete floor to eliminate the treadmill's lowest factory vibration mode
    (~5 hz). But these blocks seem to have corrupted the force plate
    measurements by imposing frictional stresses on the system. The force plate
    measurements should not be used from these trials, but the marker data is
    fine.
  \item We did not record unloaded compensation trials for trials 9-15.
    Regardless, they would be useless due to the previous point.
  \item Trials 6-8 use an early experimental protocol which divided the
    perturbation sections into three sections: longitudinal perturbations,
    lateral perturbations, and a combination of each. We then learned the
    treadmill had a low vibrational mode which significantly affects the force
    plate measurements. The force measurements during these trials are
    corrupted by this vibrational mode and should be used with caution or not
    at all.
  \item Trials 6-8 use a XX lower body marker set only.
  \item The ankle joint torques computed from subject 9's data in trials 25-27
    are abnormal and should be used with caution or not at all. We were not
    about to locate the source of the error, but it is likely related to the
    force calibration.
\end{itemize}
